---
title: "DV Assignment 2: Feedback Presentation"
author: "Group 2: Teodora Costache, Andrei Dunuta, Filip Gawrylczyk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

<!--
HOW TO KNIT THIS FILE

1. Open your group project in RStudio first (the `Project:` button in the
   top right should show your repo name). Keep this file in `report/`.
2. Click the `Knit` button at the top of this file (or press
   Ctrl/Cmd + Shift + K). An .html file with the same name appears next to
   this .Rmd in `report/`.
3. Knitting starts from a clean R session: it only knows what is created
   *in this file*. The setup chunk below sources scripts/00-packages.R and
   scripts/01-get-data.R, so packages and data are always loaded. The first
   knit downloads the data, so you need an internet connection.

If knitting fails:
- "could not find function": the package isn't loaded - add it to
  scripts/00-packages.R.
- "object not found": you created it in the Console, not in this file (or
  in a script that this file doesn't source).
- "cannot open file": check that the project is open, and use
  here("folder", "file") for paths instead of typing them out.
- Check that all code chunks have unique labels.
- Google the error message you receive, as it often provides clues about
  what went wrong.

This comment does not show up in the knitted .html, so you can leave it here.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))

library(tidyverse)
```

## Research question

RQ 3. Do schools with similar test results give similar secondary-school advice?

*Level*: Inference

*Target audience*: policymakers (education ministry).

## Exploratory process

*Describe how you explored the data before settling on your final visualization. Include at least one exploratory plot below (this can be the one from Part 1, or an updated one), and a short note on what it told you and how it influenced your final visualization.*


### Data preparation and choices 

We excluded schools with <5 students, because the number of students and the amount of advice would simply be unknown for these schools, making it impossible to compute the percentage of students who got specific advice.

Percentage of students that passed reference levels was computed per school as the average of the sum of tests passed divided by all tests taken, per school. Each student took 3 tests, mathematics, and 2 reading tests. Based on that, for example, if a school had 100 students, each student would take 3 tests, therefore there would be 300 tests taken. If in that school, 70 students passed math, 30 students passed reading 1, and 50 students passed reading 2, the percentage of students that passed the reference level would be 50% (70 + 30 + 50)/300 = 150/300 = 0.5 = 50%.

Since small numbers of students with each advice are not counted precisely below 5 and are given a "<5" marking, we computed the average of what these <5 mean in the dataset, which could be done thanks to the amount of students. Through this, we found that an average "<5" marking on advice equals 1.621641 students.

Our dataset for computations already did not include any special schools (Sbo). Even if it did, we decided it would be wiser to drop it, because special schools have different criteria for giving school advice.

Schools with no advice data ("NA") were dropped.

School advice has 12 possible levels in order from easiest to achieve to hardest to achieve:
VSO < PRO < VMBO-B < VMBO-B/VMBO-K < VMBO-K < VMBO-K/VMBO-GT < VMBO-GT < VMBO-GT/HAVO < HAVO < HAVO/VWO < VWO.
Whenever there is a mixed advice, for example HAVO/VWO, that means that a student has a mixed track for the first years of high school - after which one of the two possibilities is chosen for the rest of high school. 
```{r data-preparation}
# Taking this from assignment 1 and adding VESTIGINGSCODE
# as a grouper, so we can use every school.
school_provider <- eindscores |>
  # The filter below only filters out regular schools.
  filter(SOORT_PO == "Bo") |>
  select(INSTELLINGSCODE, VESTIGINGSCODE, matches("_(AANTAL|GEM)$")) |>
  pivot_longer(
    cols = matches("_(AANTAL|GEM)$"),
    names_to = c("provider", ".value"),
    names_pattern = "(.+)_(AANTAL|GEM)"
  ) |>
  mutate(AANTAL = suppressWarnings(as.numeric(AANTAL))) |>
  filter(!is.na(AANTAL), AANTAL > 0, !is.na(GEM), GEM > 0) |>
  group_by(INSTELLINGSCODE, VESTIGINGSCODE, provider) |>
  summarise(AANTAL = sum(AANTAL), .groups = "drop")
# Same as assignment 1, adding VESTIGINGSCODE.
primary_provider <- school_provider |>
  group_by(INSTELLINGSCODE, VESTIGINGSCODE) |>
  slice_max(AANTAL, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(INSTELLINGSCODE, VESTIGINGSCODE, provider)

count_cols <- c(
  "REKENEN_LAGER1F", "REKENEN_1F", "REKENEN_1S",
  "LV_LAGER1F", "LV_1F", "LV_2F",
  "TV_LAGER1F", "TV_1F", "TV_2F"
)
# Same as assignment 1 to make the score column,
# of many people "pass" both literature and math level. Also
# added VESTIGINGSCODE again.
ref_levels <- referentieniveaus |>
  mutate(across(all_of(count_cols), ~ suppressWarnings(as.numeric(.)))) |>
  group_by(INSTELLINGSCODE, VESTIGINGSCODE) |>
  summarise(
    across(all_of(count_cols), ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  mutate(
    rekenen_n = REKENEN_LAGER1F + REKENEN_1F + REKENEN_1S,
    taal_lv_n = LV_LAGER1F + LV_1F + LV_2F,
    taal_tv_n = TV_LAGER1F + TV_1F + TV_2F,
    score = (
      100 * REKENEN_1S / rekenen_n +
        100 * LV_2F / taal_lv_n +
        100 * TV_2F / taal_tv_n
    ) / 3,
    n_students = rekenen_n
  ) |>
  filter(rekenen_n > 0, taal_lv_n > 0, taal_tv_n > 0) |>
  select(INSTELLINGSCODE, VESTIGINGSCODE, score, n_students) 

# Joining reference with primary provider.
school_data <- ref_levels |>
  inner_join(primary_provider, by = c("INSTELLINGSCODE", "VESTIGINGSCODE"))

# Now to the actual changes -
# joining with schooladviezen, to have school advices
# per school.
school_data <- school_data |>
  left_join(schooladviezen |>
    select(
           INSTELLINGSCODE:INSTELLINGSNAAM_VESTIGING,
           PLAATSNAAM,
           GEMEENTENAAM,
           PROVINCIE,
           VSO:ADVIES_NIET_MOGELIJK),
  by = c("INSTELLINGSCODE", "VESTIGINGSCODE"))

# Since DUO writes small numbers as <5, the options
# are for it to be 1, 2, 3 or 4, and the average is 2.5.
# BUT THEN if we use 2.5 the numbers get inflated a lot over
# n_student, for example a school with 13 student had a total
# of advices as 30 - this happened with many schools.
# Because of this we tried to use 2 instead of 2.5 -
# close to the average of possibilities but a bit lower, but
# it was still too much overshoot.
# Later on we decided to find the exact number that would
# make an average percentage count towards 100%, the computation
# is found later as part of advice_check.


school_data <- school_data |>
  mutate(across(VSO:ADVIES_NIET_MOGELIJK,
                ~ as.numeric(replace(.x, .x == "<5", "1.621641"))))

# Dropping any NAs that might be in the data and combining
# different advice types into at least VWO, at least HAVO
# (without the VWO) and all lower than those.

school_data <- school_data |>
  drop_na(score, n_students, PRO:VWO) |>
  group_by(INSTELLINGSCODE, VESTIGINGSCODE) |>
  mutate(at_least_VWO = HAVO_VWO + VWO,
         at_least_HAVO = VMBO_GT_HAVO + HAVO,
         VMBO_or_worse = rowSums(pick(PRO:VMBO_GT)))
# Now time to check what numbers are the best to replace <5 with
# so that there is no issue with schools having too many
# or too few students. When we were using 2, this was an issue.

advice_check <- school_data |>
  group_by(INSTELLINGSCODE, VESTIGINGSCODE) |>
  mutate(total_advice = 
           at_least_VWO + at_least_HAVO + VMBO_or_worse + VSO + ADVIES_NIET_MOGELIJK,
         over_advice = total_advice - n_students) |>
  summarize(total_advice, n_students, over_advice, .groups = "drop_last")
# The mean with <5 being replaced by 2 was 1.8.
# Then we computed a number that would make it 0. IF all computations
# go correctly, all the following numbers will equal 0.

# Useful when writing the code, but it clutters the report
# mean(advice_check$over_advice)

# This is the sum by which our replacement overshoots.
sum_over <- sum(advice_check$over_advice)

# This is how many numbers replacing <5 we have in a dataset
n_two <- school_data |>
  ungroup() |>
  select(VSO:VWO) |>
  as.matrix() |>
  (\(x) sum(x == 1.621641, na.rm = TRUE))()

# Simple division to lower the overshoot by the amount we need.
# Currently should equal the correct number, because
# the overshoot is 0

# Useful when writing the code, but it clutters the report
# 1.621641 - (sum_over / n_two)

# Now we will calculate percentages of students
# in each school that got each advice.
# Later, since some schools had a lot too many or too few
# advices compared to students, we normalize the number
# into advice divided by advices given * 100 for percentage.
# At the end, filtering out one school in Utrecht that gave
# no advice at all, so along the way we divided by 0.

school_data <- school_data |>
  mutate(VWO_percent = (at_least_VWO / n_students) * 100,
         HAVO_percent = (at_least_HAVO / n_students) * 100,
         VMBO_percent = (VMBO_or_worse / n_students) * 100,
         total = VWO_percent + HAVO_percent + VMBO_percent,
         across(VWO_percent:VMBO_percent, ~ 100 * .x / total)) |>
  select(!VSO:VMBO_or_worse, -total) |>
  filter(is.finite(score) &
           is.finite(VWO_percent) &
           is.finite(HAVO_percent) &
           is.finite(VMBO_percent))

# Now we pivot the dataframe so that it is easier to put it into
# ggplot.

school_data <- school_data |>
  pivot_longer(c(VWO_percent, HAVO_percent, VMBO_percent),
               names_to = "group", names_pattern = "(.+)_percent",
               values_to = "pct") |>
  mutate(group = factor(group, levels = c("VWO", "HAVO", "VMBO")))


# In case there are multiple locations (vestings) in a school all the
# schoolweging scores in that school (instellings) are averaged. This way there
# is for sure only one score per school and it is easier to compute.

# Right now, the codes for each school/location are merged in the OVT column
# (i.e 	00AP|C1, which has the intellingscode and the vestingscode). I am first
# separating them so that I can filter the schools properly, then averagnig the
# scores.

schoolweging_by_school <- schoolweging |>
  separate(OVT, into = c("INSTELLINGSCODE", "schoolweging_location"),
           sep = "\\|", remove = FALSE) |>
  mutate(
    schoolweging = as.numeric(str_replace(as.character(schoolweging),
                                          ",", ".")),
    aantal_leerlingen = as.numeric(str_replace(as.character(aantal_leerlingen),
                                               ",", "."))
  ) |>
  filter(!is.na(INSTELLINGSCODE), !is.na(schoolweging)) |>
  group_by(INSTELLINGSCODE) |>
  summarise(
    schoolweging = weighted.mean(
      schoolweging,
      w = if_else(is.na(aantal_leerlingen) | aantal_leerlingen <= 0, 1,
                  aantal_leerlingen)
    ),
    .groups = "drop"
  )

#IM saving all the averaged schoolweging scores in a separate data set so that
# I can work with it without getting in the way of anyone else working on the
# school_data variable.
school_data_schoolweging <- school_data |>
  left_join(schoolweging_by_school, by = "INSTELLINGSCODE") |>
  filter(is.finite(schoolweging), is.finite(pct))
```


```{r exploratory-plot, fig.width = 7.5, fig.height = 10}
# Making plot with score on x axis, and each line
# representing the percentage of students with each advice.
# Y-axis is that percentage.
exp1 <- ggplot(school_data, aes(x = score, y = pct)) +
  # We add points for each school on the plot to be able
  # to visualize the spread
  geom_point(aes(color = group), alpha = 0.075) +
  # And add a line in the middle, that shows where, more or less
  # the averages are.
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"),
              se = FALSE, color = "black") +
  # Now with geom_quantile we create quantile lines, that
  # show us where 80% of schools fall, to be able
  # to see whether different schools give similar advice
  # if they have similar test results, the line is dashed.
  geom_quantile(formula = y ~ x, quantiles = c(0.1, 0.9), color = "black",
                linetype = "dashed", linewidth = 1) +
  # In order to see all 3 plots at once we use facet wrap.
  facet_wrap(~ group, nrow = 1,
             labeller = as_labeller(c(VWO = "VWO and VWO/HAVO",
                                      HAVO = "HAVO and HAVO/VMBO",
                                      VMBO = "VMBO, PRO and VSO"))) +
  # And we add some colors that are quite different from one another
  # But still show that there are factors so VMBO -> HAVO -> VWO
  scale_color_manual(values = c(VWO = "#E66101",
                                HAVO = "#9E4F9E",
                                VMBO = "#2166AC"),
                     guide = "none") +
  # Adding normal labels.
  labs(x = "% of students reaching target level, per school",
       y = "% of advice given, per school",
       title = paste0("Do schools with similar target level ",
                      "scores give similar school advice?"),
       subtitle = paste0("Solid lines show average trend. Dashed lines ",
                         "show where 80% of schools fall")) +
  # Since the percentage chance is 0 to 100 for both axes,
  # we changed both axes to show the entire range of possibilities.
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  # theme_minimal with larger text size and no minor grid.
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())



# This graph looks at whether schools with a similar schoolweging score tend to
# give similar advice, and what that advice tends to be.

# The specific question I had in mind was: Do schools with a higher schooling
# score (more disadvantage) tend to recommand schools of a lower rank?

#This graph is stylistically the same as the one above, for the sake of
# consistancy. But I do want to make more, just to play around.

exp2 <- ggplot(school_data_schoolweging, aes(x = schoolweging, y = pct)) +
  geom_point(aes(color = group), alpha = 0.075) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"),
              se = FALSE, color = "black") +
  geom_quantile(formula = y ~ x, quantiles = c(0.1, 0.9), color = "black", 
                linetype = "dashed", linewidth = 1) +
  facet_wrap(~ group, nrow = 1,
             labeller = as_labeller(c(VWO = "VWO and HAVO/VWO",
                                      HAVO = "HAVO and HAVO/VMBO",
                                      VMBO = "VMBO, PRO and VSO"))) +
  scale_color_manual(values = c(VWO = "#E66101",
                                HAVO = "#9E4F9E",
                                VMBO = "#2166AC"), guide = "none") +

  labs(
    x = "School weighting (higher = more disadvantaged pupil population)",
    y = "% of advice given, per school",
    title = "Do schools with similar schoolweging scores give similar advice?",
    subtitle = paste0("Solid lines show average trends;",
                "dashed lines show where 80% of schools fall")
  ) +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())



plot_grid(exp1, exp2, nrow=2)
```

## Final visualization

*Your final, explanatory visualization goes below. Include a short write-up: what does it communicate, why is this the best way to communicate it, and how robust is the finding?*

```{r final-plot}
```

## Conclusion

*What can you conclude from your visualization? What are the limitations of this analysis?*

---

**Before submitting:** run the lintr check on this file, and re-knit from a clean environment (`rm(list = ls())`, already done above) to make sure everything reproduces from scratch.
