---
title: "DV Assignment 2, Part 1: [Your Title Here]"
author: "Group 2: Filip Gawrylczyk, Teodora Costache, Andrei Dunuta"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

<!--
HOW TO KNIT THIS FILE

1. Open your group project in RStudio first (the `Project:` button in the
   top right should show your repo name). Keep this file in `report/`.
2. Click the `Knit` button at the top of this file (or press
   Ctrl/Cmd + Shift + K). An .html file with the same name appears next to
   this .Rmd in `report/`.
3. Knitting starts from a clean R session: it only knows what is created
   *in this file*. The setup chunk below sources scripts/00-packages.R and
   scripts/01-get-data.R, so packages and data are always loaded. The first
   knit downloads the data, so you need an internet connection.

If knitting fails:
- "could not find function": the package isn't loaded - add it to
  scripts/00-packages.R.
- "object not found": you created it in the Console, not in this file (or
  in a script that this file doesn't source).
- "cannot open file": check that the project is open, and use
  here("folder", "file") for paths instead of typing them out.
- Check that all code chunks have unique labels.
- Google the error message you receive, as it often provides clues about
  what went wrong.

This comment does not show up in the knitted .html, so you can leave it here.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))

# Loading necessary libraries.
library(tidyverse)
```

## Research question

*What is your research question? Is it a question of analytics, inference or prediction? Who is the target audience for your final visualization, and how does that shape your choices?*

## Exploratory process

*Describe how you explored the data. Include at least one exploratory visualization below, and a short note on what it shows, and what you learned from it about your research question.*
```{r data-preparation-plot1}
# Taking this from assignment 1 and adding VESTIGINGSCODE
# as a grouper, so we can use every school.
school_provider <- eindscores |>
  select(INSTELLINGSCODE, VESTIGINGSCODE, matches("_(AANTAL|GEM)$")) |>
  pivot_longer(
    cols = matches("_(AANTAL|GEM)$"),
    names_to = c("provider", ".value"),
    names_pattern = "(.+)_(AANTAL|GEM)"
  ) |>
  # DUO writes "<5" for small counts, so we decided to get
  # rid of it because the computations would make no sense.
  mutate(AANTAL = suppressWarnings(as.numeric(AANTAL))) |>
  filter(!is.na(AANTAL), AANTAL > 0, !is.na(GEM), GEM > 0) |>
  group_by(INSTELLINGSCODE, VESTIGINGSCODE, provider) |>
  summarise(AANTAL = sum(AANTAL), .groups = "drop")
# Same as assignment 1, adding VESTIGINGSCODE.
primary_provider <- school_provider |>
  group_by(INSTELLINGSCODE, VESTIGINGSCODE) |>
  slice_max(AANTAL, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(INSTELLINGSCODE, VESTIGINGSCODE, provider)

count_cols <- c(
  "REKENEN_LAGER1F", "REKENEN_1F", "REKENEN_1S",
  "LV_LAGER1F", "LV_1F", "LV_2F",
  "TV_LAGER1F", "TV_1F", "TV_2F"
)
# Same as assignment 1 to make the score column,
# of many people "pass" both literature and math level. Also
# added VESTIGINGSCODE again.
ref_levels <- referentieniveaus |>
  mutate(across(all_of(count_cols), ~ suppressWarnings(as.numeric(.)))) |>
  group_by(INSTELLINGSCODE, VESTIGINGSCODE) |>
  summarise(
    across(all_of(count_cols), ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  mutate(
    rekenen_n = REKENEN_LAGER1F + REKENEN_1F + REKENEN_1S,
    taal_lv_n = LV_LAGER1F + LV_1F + LV_2F,
    taal_tv_n = TV_LAGER1F + TV_1F + TV_2F,
    score = (
      100 * REKENEN_1S / rekenen_n +
        100 * LV_2F / taal_lv_n +
        100 * TV_2F / taal_tv_n
    ) / 3,
    n_students = rekenen_n
  ) |>
  filter(rekenen_n > 0, taal_lv_n > 0, taal_tv_n > 0) |>
  select(INSTELLINGSCODE, VESTIGINGSCODE, score, n_students)
# Joining reference with primary provider.
school_data <- ref_levels |>
  inner_join(primary_provider, by = c("INSTELLINGSCODE", "VESTIGINGSCODE"))

# Now to the actual changes -
# joining with schooladviezen, to have school advices
# per school.
school_data <- school_data |>
  left_join(schooladviezen |>
    select(
           INSTELLINGSCODE:INSTELLINGSNAAM_VESTIGING,
           PLAATSNAAM,
           GEMEENTENAAM,
           PROVINCIE,
           VSO:ADVIES_NIET_MOGELIJK),
  by = c("INSTELLINGSCODE", "VESTIGINGSCODE"))

# Since DUO writes small numbers as <5, the options
# are for it to be 1, 2, 3 or 4, and the average is 2.5.
# BUT THEN if we use 2.5 the numbers get inflated a lot over
# n_student, for example a school with 13 student had a total
# of advices as 30 - this happened with many schools.
# Because of this we tried to use 2 instead of 2.5 -
# close to the average of possibilities but a bit lower, but
# it was still too much overshoot.
# Later on we decided to find the exact number that would
# make an average percentage count towards 100%, the computation
# is found later as part of advice_check.


school_data <- school_data |>
  mutate(across(VSO:ADVIES_NIET_MOGELIJK,
                ~ as.numeric(replace(.x, .x == "<5", "1.621641"))))

# Dropping any NAs that might be in the data and combining
# different advice types into at least VWO, at least HAVO
# (without the VWO) and all lower than those.

school_data <- school_data |>
  drop_na(score, n_students, VSO:VWO) |>
  group_by(INSTELLINGSCODE, VESTIGINGSCODE) |>
  mutate(at_least_VWO = HAVO_VWO + VWO,
         at_least_HAVO = VMBO_GT_HAVO + HAVO,
         VMBO_or_worse = rowSums(pick(VSO:VMBO_GT)))
# Now time to check what numbers are the best to replace <5 with
# so that there is no issue with schools having too many
# or too few students. When we were using 2, this was an issue.

advice_check <- school_data |>
  group_by(INSTELLINGSCODE, VESTIGINGSCODE) |>
  mutate(total_advice = at_least_VWO + at_least_HAVO + VMBO_or_worse,
         over_advice = total_advice - n_students) |>
  summarize(total_advice, n_students, over_advice, .groups = "drop_last")
# The mean with <5 being replaced by 2 was 1.8.
# Then we computed a number that would make it 0. IF all computations
# go correctly, all the following numbers will equal 0.
mean(advice_check$over_advice)

# This is the sum by which our replacement overshoots.
sum_over <- sum(advice_check$over_advice)

# This is how many numbers replacing <5 we havein a dataset
n_two <- school_data |>
  ungroup() |>
  select(VSO:VWO) |>
  as.matrix() |>
  (\(x) sum(x == 1.621641, na.rm = TRUE))()

# Simple division to lower the overshoot by the amount we need.
# Currently should equal the correct number, because
# the overshoot is 0

1.621641 - (sum_over / n_two)

# Now we will calculate percentages of students
# in each school that got each advice.
# Later, since some schools had a lot too many or too few
# advices compared to students, we normalize the number
# into advice divided by advices given * 100 for percentage.
# At the end, filtering out one school in Utrecht that gave
# no advice at all, so along the way we divided by 0.
school_data <- school_data |>
  mutate(VWO_percent = (at_least_VWO / n_students) * 100,
         HAVO_percent = (at_least_HAVO / n_students) * 100,
         VMBO_percent = (VMBO_or_worse / n_students) * 100,
         total = VWO_percent + HAVO_percent + VMBO_percent,
         across(VWO_percent:VMBO_percent, ~ 100 * .x / total)) |>
  select(!VSO:VMBO_or_worse, -total) |>
  filter(is.finite(score) &
           is.finite(VWO_percent) &
           is.finite(HAVO_percent) &
           is.finite(VMBO_percent))

# Now we pivot the dataframe so that it is easier to put it into
# ggplot.

school_data <- school_data |>
  pivot_longer(c(VWO_percent, HAVO_percent, VMBO_percent),
               names_to = "group", names_pattern = "(.+)_percent",
               values_to = "pct") |>
  mutate(group = factor(group, levels = c("VWO", "HAVO", "VMBO")))


```

```{r exploratory-plot1, fig.height = 5, fig.width = 7.5 , fig.align = "center", message=FALSE}
# Making plot with score on x axis, and each line
# representing the percentage of students with each advice.
# Y-axis is that percentage.
ggplot(school_data, aes(x = score, y = pct)) +
  # We add points for each school on the plot to be able
  # to visualize the spread
  geom_point(aes(color = group), alpha = 0.075) +
  # And add a line in the middle, that shows where, more or less
  # the averages are.
  geom_smooth(se = FALSE, color = "black") +
  # Now with geom_quantile we create quantile lines, that
  # show us where 80% of schools fall, to be able
  # to see whether different schools give similar advice
  # if they have similar test results, the line is dashed.
  geom_quantile(quantiles = c(0.1, 0.9),
                color = "black",
                linetype = "dashed",
                linewidth = 1) +
  # In order to see all 3 plots at once we use facet wrap.
  facet_wrap(~ group, nrow = 1,
             labeller = as_labeller(c(VWO = "VWO and VWO/HAVO",
                                      HAVO = "HAVO and HAVO/VMBO",
                                      VMBO = "VMBO, PRO and VSO"))) +
  # And we add some colors that are quite different from one another
  # But still show that there are factors so VMBO -> HAVO -> VWO
  scale_color_manual(values = c(VWO = "#E66101",
                                HAVO = "#9E4F9E",
                                VMBO = "#2166AC"),
                     guide = "none") +
  # Adding normal labels.
  labs(x = "% of students reaching target level, per school",
       y = "% of advice given, per school",
       title = paste0("Do schools with similar target level ",
                      "scores give similar school advice?"),
       subtitle = paste0("Solid lines show average trend. Dashed lines ",
                         "show where 80% of schools fall")) +
  # Since the percentage chance is 0 to 100 for both axes,
  # we changed both axes to show the entire range of possibilities.
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  # theme_minimal with larger text size and no minor grid.
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```
**The above plot shows % of advice given per school based on % of students in that school that reached target level. It includes lines showing that relationship for each recommendation type (like VWO or VMBO) and also includes singular schools as points. To be able to visualize the spread, and whether it is large, we showed 80% range of schools, and around where it is. It seems that there indeed is a relationship between how many children reach target level and what type of advice is given. Based on our research question, it does seem that a lot of schools, even if they have similar test results, give very different advice, which seems to be the case for all plots but especially the higher end of the VWO range, where the advice is very different between schools. For example, there are schools with 100% having reached the target. Some schools give VWO advice around >85% of the time, while other schools give VWO advice only 15% of the time.**
```{r data-preparation-plot2}
```

```{r exploratory-plot2}
```
---

**Before submitting:** run the lintr check on this file, knit it, and push your work to GitHub.
